---
title: "Getting started with PathwaySpace"
author: "Sysbiolab Team"
date: "`r Sys.Date()`"
bibliography: ../bibliography.bib
output:
  html_document:
    toc: false
---

```{css, echo=FALSE}
.card-title {
  text-align: center !important;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(bs4cards)
```

<br/>
**Package**: PathwaySpace `r packageVersion('PathwaySpace')`

For a self-contained introduction that works offline, run **vignette("PathwaySpace")** after installing the package.

# Introductory vignettes

These tutorials introduce *PathwaySpace* using simple toy examples: **Network Signal Projections** walks through the main *PathwaySpace* projection methods; **Signal Decay Functions** describes how to set up decay functions to model specific network nodes; and **Signal Aggregation Rules** demonstrates how to create signal aggregation functions to emphasize patterns in the data.

```{r, include=FALSE}
title <- c("Network Signal Projections", "Signal Decay Functions", "Signal Aggregation Rules")
image <- c("figs_intro/card1.png", "figs_intro/card2.png", "figs_intro/card3.png")
link <- c("projection-methods.html","modeling-signal-decay.html","signal-aggregation-rules.html")
ps_cards1 <- data.frame(title = title, image = image, link = link)
```

```{r, echo=FALSE}
ps_cards1 %>% 
  cards(
      title = title,
      image = image,
      link = link,
      layout = "label-below", 
      width = 3,
      border_radius = 5
  )
```

---

<div class="mt-5 mb-4"></div>

# Visualizing features in *PathwaySpace*

These tutorials showcase *PathwaySpace* applied to real-world scenarios, highlighting complex feature patterns. The **Sparse Feature Sets** vignette demonstrates how to analyze feature signals when only a limited number of nodes carry relevant information, while the **Spatial Transcriptomics** vignette illustrates how to project gene expression signals across tissue sections.

```{r, include=FALSE}
title <- c("Sparse Feature Sets", "Spatial Transcriptomics")
image <- c("figs_fsets/card1.png", "figs_spatl/card1.png")
link <- c("sparse-feature-sets.html","spatial-transcriptomics.html")
ps_cards2 <- data.frame(title = title, image = image, link = link)
```

```{r, echo=FALSE}
ps_cards2 %>% 
  cards(
      title = title,
      image = image,
      link = link,
      layout = "label-below", 
      width = 3,
      border_radius = 5
  )
```



